---
title: "2025 Per Capita Peachtree Lottery Sales"
date: "June 01, 2026"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

The Peachtree Lottery is a $2.4 billion dollar industry that raises *$650 million annually* for Georgia scholarships. We want to market to potential lottery players in a way that maximizes revenue while promoting responsible play. This report examines who bought tickets in 2025, how much they spent, and where they are located.

```{r}
#| label: setup

```

## Geographic Distribution

The map below shows lottery spending across Georgia counties. Darker colors indicate higher per-capita spending.

```{r}
#| label: map
#| fig-height: 6


```

## Income distribution

The chart below explores per-capita sales by income group.

```{r}
#| label: income


```

## Detailed Data Table

The table below provides the complete sales dataset, which can be searched interactively.

```{r}
#| label: data-table

```

---

*Data source: Simulated based on Georgia Lottery sales*