---
title: "2025 Cancer Incidence Rates in Georgia"
format: html
execute:
  echo: false
  warning: false
  message: false
---

<!-- Note: Move this file into your main directory before previewing. The filepaths used below will not work if you keep it in the solutions/ directory. -->

## Overview

Cancer incidence varies significantly across Georgia counties and demographic groups. This report examines diagnosed cancer cases in 2025, exploring how rates differ by race/ethnicity and where cases are concentrated geographically.

```{r}
#| label: setup

library(tidyverse)
library(leaflet)
library(sf)
library(tigris)
library(DT)

# load data
georgia_cases <- readRDS("georgia_cases.RDS")

# load mapping function
source("helpers.R")
```

## Geographic Distribution

The map below shows cancer incidence rates across Georgia counties. Darker colors indicate higher rates per 100,000 population.

```{r}
#| label: map
#| fig-height: 6

create_incidence_map(georgia_cases, rate = TRUE)
```

## Incidence by Race/Ethnicity

The chart below explores cancer incidence rates by race/ethnicity group.

```{r}
#| label: income

plot_incidence_by_demographic(georgia_cases, RE)
```

## Detailed Data Table

The table below provides the complete cancer cases dataset, which can be searched interactively.

```{r}
#| label: data-table

datatable(georgia_cases)
```

---

*Data source: Simulated based on Georgia cancer incidence data*
