---
title: "2025 Per Capita Peachtree Lottery Sales"
date: "June 01, 2026"
format: html
execute:
  echo: false
  warning: false
  message: false
---

<!-- Note: Move this file into your main directory before previewing. The filepaths used below will not work if you keep it in the solutions/ directory. -->

## Overview

The Peachtree Lottery is a $2.4 billion dollar industry that raises *$650 million annually* for Georgia scholarships. We want to market to potential lottery players in a way that maximizes revenue while promoting responsible play. This report examines who bought tickets in 2025, how much they spent, and where they are located.

```{r}
#| label: setup

library(tidyverse)
library(leaflet)
library(sf)
library(tigris)
library(DT)

# load data
peachtree_lottery <- readRDS("peachtree_lottery.RDS")

# load mapping function
source("lottery_helpers.R")
```

## Geographic Distribution

The map below shows lottery spending across Georgia counties. Darker colors indicate higher per-capita spending.

```{r}
#| label: map
#| fig-height: 6

create_lottery_map(peachtree_lottery, per_capita = TRUE)
```

## Demographic distribution

The charts below explore per-capita sales by age group, sex, and income level.

:::{layout-ncol="3"}
```{r}
#| label: age_group
#| fig-height: 3
#| fig-width: 2

plot_per_capita_spend(age, color = "#66C2A5")
```

```{r}
#| label: sex
#| fig-height: 3
#| fig-width: 2

plot_per_capita_spend(sex, color = "#8DA0CB")
```

```{r}
#| label: income
#| fig-height: 3
#| fig-width: 2

plot_per_capita_spend(income, color = "#E78AC3")
```
:::

## Detailed Data Table

The table below provides the complete sales dataset, which can be searched interactively.

```{r}
#| label: data-table

datatable(peachtree_lottery)
```

---

*Data source: Simulated based on Georgia Lottery sales*